Metropolitan Statistical Area Evaluation

### 2022-11-11 src/demo_data_wrangling.Rmd
calculating normalized and employment rates

# create __ 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% 
  #separate(col = "state", into = c("state1", "state2", "state3"),    sep = "-") %>% 
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by Owen 
  select(cbsa, MSA, state, pop_percent_change20_21)

@@

# join population growth data set to household median income
MSA_house_income <- MSA_income_data %>% # rename row names
  inner_join(potential_MSAs) %>% 
  select(cbsa, MSA, state,median_income, income_moe_USD) 
#select(cbsa, MSA, state, pop_percent_change20_21, median_income, income_moe_USD)

@@

```{r, household income visual, fig.height = 7, fig.align='center'}
ggplot(MSA_house_income, aes(x = median_income, y = reorder(MSA, median_income))) +
  geom_errorbarh(aes(xmin = median_income - income_moe_USD, xmax = median_income + income_moe_USD)) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median household income in 2021", 
       subtitle = "", 
       x = "estimate (in USD)", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2015) %>% 
  full_join(MSA_house_income) %>% 
  drop_na() %>% 
  select(cbsa, MSA, year, month, employment_rate, median_income, income_moe_USD)

# calculate YoY growth 
MSA_employment_yoy <- MSA_employment %>% 
  filter(year >= 2016) %>% 
  filter(month == "01") %>% 
  arrange(MSA, month) %>% 
  group_by(MSA) %>% 
  mutate(YoY = (employment_rate-lag(employment_rate))/lag(employment_rate)*100) %>% 
  mutate(employment_YoY = round(YoY, digits = 3)) %>% 
  mutate(emp_pos = employment_YoY >= 0) %>% # function to create column for yoy value, true = positive YoY
  select(!c(YoY, month)) %>% 
  drop_na()

@@

@@

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10}
# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment_yoy, aes(x = year, y = employment_rate, color = MSA)) +
  geom_line() +
  xlim(2016, 2022) +
  theme(legend.position = "none") +

@@

```{r payroll data}
# payroll data has not been added to final table
payroll_data <- read_csv(here("data","raw", "employee_payroll_MSA21_22.csv")) %>% 
  row_to_names(4) %>% 
  clean_names() %>% 
  rename(MSA = na) %>% 
  rename(payroll_change_may21 = may_2021) %>% 
  rename(payroll_change_mar22 = mar_2022) %>% 
  rename(payroll_change_apr22 = apr_2022) %>% 
  rename(payroll_change_may22 = may_2022_p)

@@

```
```{r payroll visual, fig.height = 7}
ggplot(MSA_payroll, aes(x = payroll_change, y = reorder(MSA, payroll_change), fill = payroll_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  labs(x = "percent change",
       y = "",
       title = "Quarterly change in employee payroll from 2021 to 2022") +
  facet_grid(~year)

@@

```{r}
# median rents from hud
MSA_median_rent2021_data <- read_csv(here("data", "raw", "2021MSA_medianrent.csv")) %>% 
  separate(col = cbsasub21,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("cbsa", "unknown"),
           sep = -5) %>% 
  mutate(cbsa = as.double(cbsa)) %>% 
  select(cbsa, rent50_0, rent50_1, rent50_2, rent50_3, rent50_4) %>% 
  right_join(potential_MSAs) 
```

@@

```{r, house price visual, fig.height = 7, fig.align='center'}
ggplot(metro_house_prices, aes(x = houseprice_change21_22, y = reorder(MSA, houseprice_change21_22))) +
  geom_col(fill = "dodgerblue4") +
  geom_text(aes(label = houseprice_change21_22),
            size = 4,
            hujust = 1, nudge_x = -1,
            color = "white") +
  theme_minimal()  +
  labs(x = "% change",
       y = "",
       title = "Change in House Prices in 2021")

@@

```{r property taxes data wrangling}
property_tax_data <- read_csv(here("data","raw", "wide_property_tax_data.csv"))

@@

```{r property tax visual, fig.height = 7, fig.align='center'}
ggplot(MSA_property_tax, aes(x = median_tax_USD, y = reorder(MSA, median_tax_USD))) +
  geom_errorbarh(aes(xmin = median_tax_USD - tax_moe_USD, xmax = median_tax_USD + tax_moe_USD)) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median property tax in 2019", 
       subtitle = "", 
       x = "Median Property Tax estimate (in USD)", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())
```

```{r join population, unemployment, income, and house price data}
# join employment data with income, house prices, tax, and population data
MSApop_income_housecost_tax_emp <- MSA_employment_yoy %>%
  filter(year == "2022") %>% 
  mutate(employment_rate = round(employment_rate, digits = 3)) %>% 
  select(cbsa, year, employment_rate, employment_YoY) %>% 
  right_join(MSA_property_tax) %>% 
  select(!year)

>>>>

# create __ 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% 
  #separate(col = "state", into = c("state1", "state2", "state3"),    sep = "-") %>% 
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by Owen 
  mutate(normalized_pop2021 = (popestimate2021-min(popestimate2021))/ (max(popestimate2021)-min(popestimate2021))) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021)

@@

# join population growth data set to household median income
income_potential_MSAs <- MSA_income_data %>% # rename row names
  select(cbsa, median_income) %>% 
  inner_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, median_income, pop_percent_change20_21, normalized_pop2021) #reorder columns
#select(cbsa, MSA, state, pop_percent_change20_21, median_income, income_moe_USD)

@@

```{r, household income visual, fig.height = 7, fig.align='center', eval = FALSE}
ggplot(MSA_house_income, aes(x = median_income, y = reorder(MSA, median_income))) +
  geom_errorbarh(aes(xmin = median_income - income_moe_USD, xmax = median_income + income_moe_USD)) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median household income in 2021", 
       subtitle = "", 
       x = "estimate (in USD)", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2020) %>% 
  full_join(income_potential_MSAs) %>% 
  drop_na() %>% 
  select(cbsa, MSA, year, month, employment_rate)

avg_emp_yoy <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  drop_na()

# calculate YoY growth 
MSA_employment_yoy <- MSA_employment %>% 
  filter(year >= 2020) %>% 
  filter(month == "01") %>% 
  arrange(MSA, month) %>% 
  group_by(MSA) %>% 
  mutate(YoY = (employment_rate-lag(employment_rate))/lag(employment_rate)*100) %>% 
  mutate(employment_YoY = round(YoY, digits = 3)) %>% 
  mutate(emp_pos = employment_YoY >= 0) %>%# function to create column for yoy value, true = positive YoY
  inner_join(avg_emp_yoy) %>% 
  filter(year == "2022") %>% 
  select(!c(YoY, month, year))

@@

potential_MSAs <- potential_MSAs %>%

@@

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10}
# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment_yoy, aes(x = year, y = employment_rate, color = MSA)) +
  geom_line() +
  xlim(2021, 2022) +
  theme(legend.position = "none") +

@@

```{r payroll data, eval = FALSE}
# payroll data has not been added to final table
payroll_data <- read_csv(here("data","raw", "employee_payroll_MSA21_22.csv")) %>% 
  row_to_names(4) %>% 
  clean_names() %>% 
  rename(MSA = na) %>% 
  rename(payroll_change_may21 = may_2021) %>% 
  rename(payroll_change_mar22 = mar_2022) %>% 
  rename(payroll_change_apr22 = apr_2022) %>% 
  rename(payroll_change_may22 = may_2022_p)

@@

```
```{r payroll visual, fig.height = 7, eval = FALSE}
ggplot(MSA_payroll, aes(x = payroll_change, y = reorder(MSA, payroll_change), fill = payroll_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  labs(x = "percent change",
       y = "",
       title = "Quarterly change in employee payroll from 2021 to 2022") +
  facet_grid(~year)

@@

```{r median rent}
# median rents from hud
MSA_median_rent2021_data <- read_csv(here("data", "raw", "2021MSA_medianrent.csv")) %>% 
  separate(col = cbsasub21,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = as.double(cbsa)) %>% 
  mutate(median_rent2021 = as.numeric(rent50_0)) %>% 
  select(cbsa, median_rent2021) %>% 
  right_join(potential_MSAs)

MSA_median_rent2022_data <- read_csv(here("data", "raw", "2022MSA_median_rent.csv")) %>% 
  separate(col = cbsasub22,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = as.double(cbsa)) %>% 
  mutate(median_rent2022 = as.numeric(rent_50_0)) %>% 
  select(cbsa, median_rent2022)

potential_MSAs <- potential_MSAs %>% 
  right_join(MSA_median_rent2022_data) %>% 
  drop_na()

@@

```{r, house price visual, fig.height = 7, fig.align='center', eval = FALSE}
ggplot(metro_house_prices, aes(x = houseprice_change21_22, y = reorder(MSA, houseprice_change21_22))) +
  geom_col(fill = "dodgerblue4") +
  geom_text(aes(label = houseprice_change21_22),
            size = 4,
            hujust = 1, nudge_x = -1,
            color = "white") +
  theme_minimal()  +
  labs(x = "% change",
       y = "",
       title = "Change in House Prices in 2021")

@@

```{r property taxes data wrangling, eval = FALSE}
property_tax_data <- read_csv(here("data","raw", "wide_property_tax_data.csv"))

@@

```{r property tax visual, fig.height = 7, fig.align='center', eval = FALSE}
ggplot(MSA_property_tax, aes(x = median_tax_USD, y = reorder(MSA, median_tax_USD))) +
  geom_errorbarh(aes(xmin = median_tax_USD - tax_moe_USD, xmax = median_tax_USD + tax_moe_USD)) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5) + 
  labs(title = "Median property tax in 2019", 
       subtitle = "", 
       x = "Median Property Tax estimate (in USD)", 
       y = "") + 
  scale_x_continuous(labels = label_dollar())
```

```{r join population, unemployment, income, and house price data}
# join employment data with income, house prices, tax, and population data
MSApop_income_housecost_tax_emp <- MSA_employment_yoy %>%
  filter(year == "2022") %>% 
  mutate(employment_rate = round(employment_rate, digits = 3)) %>% 
  select(cbsa, year, employment_rate, employment_YoY) %>% 
  # right_join(MSA_property_tax) %>% 
  select(!year)
